Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system with plausible mechanism

### 2024-07-30 _tex/index.tex
Built site for gh-pages

In statistics, one application that incorporates concepts from graph
theory is causal inference. Specifically, the structural approach to
causal inference uses directed acyclic graphs (DAG) to provide a formal
and graphical representation of a system's causal structure
\citep{Neal_2020}. In this context, a \emph{graph} is a collection of
nodes connected by edges, where nodes represent random variables. The
term \emph{directed} indicates that the edges of the graph extend from a
\emph{parent} node to a \emph{child} node, with arrows showing the
direction of causal influence. The term \emph{acyclic} indicates that
the direction of causal influences does not form a loop, ensuring that
the influences do not cycle back on themselves \citep{McElreath_2020}.

A key advantage of DAGs is that, regardless of the complexity of the
causal structure, these structures can be composed using a minimal set
of building blocks. Moreover, representing any complex structure
requires only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. These advantages imply that a
researcher seeking to depict and understand the causal assumptions of a
complex system only needs to decompose the structure into these basic
building blocks and then analyze the causal assumptions associated with
each block \citep{McElreath_2020}.

Figure~\ref{fig-dags} presents the five fundamental building blocks of
any DAG. Figure~\ref{fig-dag_bb1} illustrates two unconnected nodes,
representing an scenario where variables \(X_{1}\) and \(X_{2}\) are not
causally related. Figure~\ref{fig-dag_bb2} shows two connected nodes,
depicting an scenario where a \emph{parent} node \(X_{1}\) has a causal
influence on a \emph{child} node \(X_{2}\). This implies that variable
\(X_{2}\) is a \emph{descendant} of \(X_{1}\). Figure~\ref{fig-dag_bb3}
illustrates a \emph{chain} (or \emph{pipe}), where \(X_{1}\) influences
\(X_{2}\), and \(X_{2}\) influences \(X_{3}\). This DAG indicates that
\(X_{1}\) is a parent node of \(X_{2}\), and \(X_{2}\) is a parent node
\(X_{3}\). Furthermore, it shows that \(X_{1}\) is an \emph{ancestor} of
\(X_{3}\), with the relationship between the two variables being
mediated entirely by \(X_{2}\). Figure~\ref{fig-dag_bb4} depicts a
\emph{fork}, illustrating an scenario where variables \(X_{1}\) and
\(X_{3}\) are both influenced by \(X_{2}\). In this scenario \(X_{2}\)
is a parent node for both \(X_{1}\) and \(X_{3}\). Finally,
Figure~\ref{fig-dag_bb5} illustrates a \emph{collider} (also known as
\emph{inmorality}), where variables \(X_{1}\) and \(X_{3}\) are
concurrent causes of \(X_{2}\). In this configuration, \(X_{1}\) and
\(X_{3}\) are not causally related to each other but both influence
\(X_{2}\). Given the heuristic nature of DAGs, the use of fundamental
building block to construct a causal structure of a system is easier to
understand using a motivating example. The motivating example can also
serve to point out about other conventions when using DAGs.

@@

Consider a system where variables \(X\) and \(Z\) influence a third
variable \(Y\). In this system, it is assumed that \(X\) and \(Z\) are
dependent on their own processes and are, therefore, independent from
each other. Figure~\ref{fig-dag_example1} presents the plausible causal
structure of this system. The DAG shows the endogenous variables
\(V=\{X,Z,Y\}\) as circled black nodes, indicating these variables are
observed. The arrows connecting the variables indicate the direction of
causal influence, while a lack of influence is indicated by the absence
of arrows. Moreover, the exogenous variables \(E=\{e_{X},e_{Z},e_{Y}\}\)
represent everything else that is chosen not to be modeled explicitly.
These exogenous variables, or \emph{disturbances}, are usually
represented by open circles to indicate their unobserved nature.
Although this DAG explicitly shows the exogenous variables,
conventionally these are omitted for brevity, resulting in an equivalent
graph as shown in Figure~\ref{fig-dag_example2}.

>>>>

In statistics, one application incorporating concepts from graph theory
is causal inference. Specifically, the structural approach to causal
inference uses directed acyclic graphs (DAG) to provide a formal and
graphical representation of the causal structure of a system
\citep{Neal_2020}. In this context, a \emph{graph} is a collection of
nodes connected by edges, where nodes represent random variables. The
term \emph{directed} indicates that the edges of the graph extend from
one node to another, with arrows showing the direction of causal
influence. Moreover, the term \emph{acyclic} indicates the causal
influences do not form a loop, meaning the influences do not cycle back
on themselves \citep{McElreath_2020}.

Regardless of complexity, DAGs can represent various causal structures
using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. Each panel of Figure~\ref{fig-dags}
illustrates these building blocks. Figure~\ref{fig-dag_bb1} depicts two
unconnected nodes, representing an scenario where variables \(X_{1}\)
and \(X_{2}\) are not causally related. Figure~\ref{fig-dag_bb2} shows
two connected nodes, illustrating a scenario where a parent node
\(X_{1}\) exerts a causal influence on a child node \(X_{2}\).
Consequently, \(X_{2}\) is considered a \emph{descendant} of \(X_{1}\).
Figure~\ref{fig-dag_bb3} depicts a \emph{chain} (or \emph{pipe}), where
\(X_{1}\) influences \(X_{2}\), and \(X_{2}\) influences \(X_{3}\). In
this configuration, \(X_{1}\) is a parent node of \(X_{2}\), and
\(X_{2}\) a parent node of \(X_{3}\). Furthermore, the DAG shows that
\(X_{1}\) is an \emph{ancestor} of \(X_{3}\), and that the relationship
between these variables is entirely mediated by \(X_{2}\).
Figure~\ref{fig-dag_bb4} illustrates a \emph{fork}, where variables
\(X_{1}\) and \(X_{3}\) are both influenced by \(X_{2}\). In this
scenario \(X_{2}\) is a parent node of both \(X_{1}\) and \(X_{3}\).
Finally, Figure~\ref{fig-dag_bb5} depicts a \emph{collider}, also known
as \emph{inmorality}, where variables \(X_{1}\) and \(X_{3}\) are
concurrent causes of \(X_{2}\). In this configuration, \(X_{1}\) and
\(X_{3}\) are not causally related to each other but both influence
\(X_{2}\).

@@

Given the heuristic nature of DAGs, the use of fundamental building
block to construct a causal structure of a system is easier to
understand using a motivating example. The motivating example can also
serve to signal about other conventions when using DAGs. Consider a
system where variables \(X\) and \(Z\) influence a third variable \(Y\).
In this system, it is assumed that \(X\) and \(Z\) are dependent on
their own processes and are, therefore, independent from each other.
Figure~\ref{fig-dag_example1} presents the plausible causal structure of
this system. The DAG shows the endogenous variables \(V=\{X,Z,Y\}\) as
circled black nodes, indicating these variables are observed. The arrows
connecting the variables indicate the direction of causal influence,
while a lack of influence is indicated by the absence of arrows.
Moreover, the exogenous variables \(E=\{e_{X},e_{Z},e_{Y}\}\) represent
everything else that is chosen not to be modeled explicitly. These
exogenous variables, or \emph{disturbances}, are usually represented by
open circles to indicate their unobserved nature. Although this DAG
explicitly shows the exogenous variables, conventionally these are
omitted for brevity, resulting in an equivalent graph as shown in
Figure~\ref{fig-dag_example2}.
